Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-08-30 manuscript/manuscript.tex
simon edits

Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high-dimensional MD simulation datasets.
In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states that adhere to the Markov property.
The Markov property asserts that the dynamics in the state space is ``memoryless'':~in other words, the probability of transitioning from any state $i$ to any other state $j$ is independent of the history of the system before it was in $i$.
In order to obtain Markovian dynamics, a lag time must be chosen at which the transitions between states are analyzed.
%This lag time must be sufficiently short to resolve the dynamics of interest, but long enough such that the Markovian approximation is appropriate.

In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
Then, the count matrix is converted to a row-stochastic transition probability matrix, which is defined for the specified lag time.
To ensure that the transition probablity matrix has desirable mathematical properties, detailed balance is enforced when converting the count matrix to the transition probability matrix, which requires that,

@@

\noindent{}where $p_t(i)$ represents the probability of being in state $i$ at time $t$, and $p(i \rightarrow j)$ is the probability of transitioning from state $i$ to $j$ at the next time step.
% Detailed balance requires that the probability of transitioning from state $i$ to state $j$, conditioned upon the system being in state $i$, is the same as the probability of transitioning from state $j$ to state $i$ conditioned upon the system being in state $j$.
The requirement of detailed balance indicates that we assume our system is reversible.
We additionally require that the system is ergodic.

When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space, but short enough to resolve the dynamics in which we are interested.
Plotting the implied timescales (ITS) as a function of $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via,
\begin{equation}
\label{eq:its}
t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
\end{equation}

>>>>

Markov state modeling is a mathematical framework for the analysis of time-series data, often but not limited to high-dimensional MD simulation datasets.
In its standard formulation, the creation of a Markov state model involves decomposing the phase or configuration space occupied by a system into a set of disjoint, discrete states that adhere to the Markov property.
The Markov property asserts that the dynamics in the state space is ``memoryless'':~in other words, the probability of transitioning from any state $i$ to any other state $j$ after a time $\tau$ is independent of the history of the system before it was in $i$.
%This lag time must be sufficiently short to resolve the dynamics of interest, but long enough such that the Markovian approximation is appropriate.

In order to create a Markov state model for a dynamical system, each data point in the time series is assigned to a state.
Given an appropriate lag time, every pairwise transition at that lag time is counted and stored in a count matrix.
Then, the count matrix is converted to a row-stochastic transition probability matrix, which is defined for the specified lag time.
To ensure that the transition probability matrix has desirable mathematical properties, detailed balance is enforced when converting the count matrix to the transition probability matrix, which requires that,

@@

\noindent{}where $p_t(i)$ represents the probability of being in state $i$ at time $t$, and $p(i \rightarrow j)$ is the probability of transitioning from state $i$ to $j$ at the next time step.
% Detailed balance requires that the probability of transitioning from state $i$ to state $j$, conditioned upon the system being in state $i$, is the same as the probability of transitioning from state $j$ to state $i$ conditioned upon the system being in state $j$.
The requirement of detailed balance indicates that we assume our system is reversible.
We additionally require that the system is ergodic, which means that every state is accessible from every other state.

When estimating an MSM it is critical to choose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our state space, but short enough to resolve the dynamics in which we are interested.
Plotting the implied timescales (ITS) as a function of $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}.
The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via,
\begin{equation}
\label{eq:its}
t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}.
\end{equation}
